---
title: "Instruction on circular regression with INLA"
output: pdf_document
date: "2026-01-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Preliminaries

```{r}
# Load required packages
require(circular)
require(INLA)
require(dplyr)
require(mvtnorm)
require(ggplot2)
require(patchwork)
```

```{r}
# Load required functions
source("../functions.R")
source("../vm.R")             # functions for von Mises distribution
source("../lavm.R")           # functions for link adjusted von Mises distribution
```

### Simulation

```{r}
# Set up model parameters
n <- 300
a0 <- 0.2
a1 <- 0.7
kappa <- 60
b0 <- 2
b1 <- 0.5
tau <- 300
```

```{r}
# Define the prior distribution
# We use the PC prior for concentration parameter of von Mises distritbuion
# Define by:
# the probability that the mean resultant length exceeds "u", is "alpha"
rprior_x <- INLA::inla.rprior.define(rprior.func, u = 0.5, alpha = 0.5)
```

```{r}
# Simulate the covariate z
z <- rnorm(n, mean = 0, sd = 1)
# As an example,
# we assume w is a second order random walk structured latent Gaussian process
# Define the precision matrix for w
Q <- INLA:::inla.rw2(n = n, sparse = TRUE, scale.model = TRUE, cyclic=!TRUE)
diag(Q) <- diag(Q) + 1e-5
# Simulate latent Gaussian process w
w <- as.numeric(
  inla.qsample(1,Q, constr=list(A=matrix(rbind(1, 1:n),2,n),e=c(0,0)))
  )
# The linear predictor for circular x
eta.x <- a0 + a1*w
# Simulate x from LAVM distribution
x <- rlavm(n, eta = eta.x, kappa = kappa)
# Simulate y from Gaussian distribution
y <- rnorm(n, mean = b0 + b1 * eta.x, sd = 1 / sqrt(tau))
```

### Bayesian inference with INLA

We first define the model response in matrix form:
```{r}
Y <- matrix(NA, 3 * n, 3)
Y[1:n,             1] <- x      # x observations (von Mises)
Y[n + 1:n,         2] <- 0      # dummy 0 for link of x
Y[2*n + 1:n,       3] <- y      # y observations (Gaussian)
```
Then, we define the covariate values and the indices of the latent processes in the linear predictor of each models:
```{r}
data_stack <- dplyr::bind_rows(
  # Rows for x (circular outcome)
  data.frame(a0    = 1,          w    = seq_len(n),
             u     = NA_real_,   e    = NA_real_,
             b0    = NA_real_,   b1   = NA_real_),
  # Rows for x link (dummy Gaussian linking latent linear predictor)
  data.frame(a0    = 1,          w    = seq_len(n),
             u     = seq_len(n), e    = -1,
             b0    = NA_real_,   b1   = NA_real_),
  # Rows for y (Gaussian outcome)
  data.frame(a0    = NA_real_,   w    = NA_real_,
             u     = NA_real_,   e    = NA_real_,
             b0    = 1,          b1   = seq_len(n))
)
```
Notice that the responses with "0" entries carry the entire linear predictor of $x$ into the linear predictor of $y$ as a whole.

Now, we define the formula to be estimated by INLA
```{r}
formula <- Y ~ -1 + a0 + b0 +
  f(w, model = "rw2", values = 1:n, scale.model = TRUE, constr = TRUE,
    hyper = list(prec = list(initial = 4,
                             prior   = "pc.prec",
                             param   = c(0.5, 0.5),
                             fixed   = FALSE))) +
  f(u, e, model = "iid", values = 1:n,
    hyper = list(prec = list(initial = -12, fixed = TRUE))) +
  f(b1, copy = "u",
    hyper = list(beta = list(initial = 0, param = c(0, 1), fixed = FALSE)))
```

We then fit the model with INLA:
```{r}
r <- inla(formula,
          data = data_stack,
          family = c("lavm", "gaussian", "gaussian"),
          control.family = list(
            list(link = "tan", 
                 hyper = list(prec = list(initial = 6,
                                          prior   = rprior_x,
                                          param   = numeric(0),
                                          fixed   = FALSE))),
            list(hyper = list(prec = list(initial = 12, fixed = TRUE))),
            list(hyper = list(prec = list(initial = 6,
                                          prior   = "pc.prec", 
                                          param   = c(0.5, 0.5),
                                          fixed   = FALSE)))
          ),
          control.inla = list(cmin = 0, compute.initial.values = TRUE),
          control.compute = list(cpo = TRUE, dic = TRUE, waic = TRUE,
                                 config = TRUE),
          verbose = FALSE)
```

We can always rerun the INLA model starting from the previous estimates to make sure the numerical optimisation is stable and more accurate. This is optional.
```{r}
r <- inla.rerun(r)
```

We can briefly skim over the INLA results with
```{r}
summary(r)
```

The entire posterior summary for the fixed effect (intercepts and coefficents for the covariates) can be check by
```{r}
r$summary.fixed
```
The posterior summary for the hyper parameters (such as model hyperparameters and coefficients related to the latent processes) can be check by
```{r}
r$summary.hyperpar
```

INLA calculate some hyperparameters in different scaling (such as log scale). To directly see the summary in intercal scale, check:
```{r}
r$internal.summary.hyperpar
```

To obtain the entire posterior marginals, we can do
```{r}
par(mfrow=c(1,2))
plot(r$marginals.fixed$a0, type = "l", xlab = expression(a[0]), ylab = "dens")
plot(r$marginals.fixed$b0, type = "l", xlab = expression(b[0]), ylab = "dens")
```
and
```{r}
par(mfrow=c(2,3))
plot(r$marginals.hyperpar$`Precision for link adjusted von Mises observations`,
     type = "l", xlab = expression(kappa), ylab = "dens")
plot(r$internal.marginals.hyperpar$`Log precision for link adjusted von Mises observations`,
     type = "l", xlab = expression(log(kappa)), ylab = "dens")
plot(r$marginals.hyperpar$`Precision for w`,
     type = "l",xlab = expression(tau[w]), ylab = "dens")
plot(r$marginals.hyperpar$`Precision for the Gaussian observations[3]`,
     type = "l", xlab = expression(tau), ylab = "dens")
plot(r$internal.marginals.hyperpar$`Log precision for the Gaussian observations[3]`,
     type = "l", xlab = expression(log(tau)), ylab = "dens")
plot(r$marginals.hyperpar$`Beta for b1`,
     type = "l", xlab = expression(b[1]), ylab = "dens")
```

Moreover, we could also obtain the fitted values for $x$ and $y$:
```{r}
par(mfrow=c(1,2))
plot(x)
lines(r$summary.fitted.values$mean[1:n], col="red", lwd=2)
plot(y)
lines(r$summary.fitted.values$mean[2*n + 1:n], col="red", lwd=2)
```
We can also obtain the posterior predictive distributions for $x$ and $y$:
```{r}
extract_draws_joint <- function(samps, idx.x, idx.y) {
  x_pred <- sapply(samps, function(s) {
    eta_vec <- s$latent[idx.x, 1]
    kappa_val <- s$hyperpar[[1]] 
    rlavm(length(eta_vec), eta = eta_vec, kappa = kappa_val)
  })
  y_pred <- sapply(samps, function(s) {
    mu <- s$latent[idx.y, 1]
    tau_val <- s$hyperpar[[length(s$hyperpar)]]
    rnorm(length(mu), mean = mu, sd = 1/sqrt(tau_val))
  })
  list(x_pred = x_pred, y_pred = y_pred)
}

n.samp <- 300
samps <- inla.posterior.sample(n.samp, r)

idx.x <- 1:n
idx.y <- 2*n + 1:n

draws <- extract_draws_joint(samps, idx.x, idx.y)
x_pred_mean <- rowMeans(draws$x_pred)
y_pred_mean <- rowMeans(draws$y_pred)

df_x <- data.frame(
  value = c(as.numeric(x), x_pred_mean),
  group = rep(c("True", "Posterior predictive"), each = length(x))
)

df_y <- data.frame(
  value = c(as.numeric(y), y_pred_mean),
  group = rep(c("True", "Posterior predictive"), each = length(y))
)

cb_palette <- c("True" = "#E69F00", "Posterior predictive" = "#56B4E9")
lineWidths <- c("True" = 1, "Posterior predictive" = 2.5)
lineTypes  <- c("True" = "solid", "Posterior predictive" = "dashed")

make_hist_styled <- function(df, xlabel) {
  ggplot(df, aes(x = value)) +
    geom_histogram(aes(y = after_stat(density), fill = group),
                   bins = 30, alpha = 0.4, position = "identity",
                   color = "white", linewidth = 0.2) +
    geom_density(aes(color = group, linewidth = group, linetype = group), alpha = 0) +
    scale_color_manual(values = cb_palette) +
    scale_fill_manual(values = cb_palette) +
    scale_linewidth_manual(values = lineWidths) +
    scale_linetype_manual(values = lineTypes) +
    labs(x = xlabel, y = NULL) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      legend.text = element_text(size = 12, face = "bold"),
      legend.key.width = unit(4, "cm"),
      panel.grid.minor = element_blank(),
      legend.margin = margin(t = 0, b = -5, unit = "pt"),
      legend.box.spacing = unit(5, "pt"),
      plot.margin = margin(5, 5, 5, 5)
    )
}

hx <- make_hist_styled(df_x, "x")
hy <- make_hist_styled(df_y, "y")

plot_predictives <- (hx | hy) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 12, face = "bold"),
    legend.key.width = unit(4, "cm"),
    legend.margin = margin(t = 0, b = -5, unit = "pt"),
    legend.box.spacing = unit(5, "pt"),
    plot.margin = margin(5, 5, 5, 5)
  )

print(plot_predictives)
```















